### 2020-01-10 ISSTA/introduction.tex
shrink

Determinism is particularly important for testing and debugging, where
being able to exactly reproduce system behavior is essential to
productivity \cite{Firesmith}.  If executing ``the same test'' can produce
significantly different behavior each time it is run, the consequences
can be unfortunate.  Developers using a test exhibiting nondeterminism
to debug a system
face a challenge in reasoning about causality, in that an event
observed may not even take place the next time the test is run; this
pernicious phenomenon has been sometimes referred to as the dreaded
``Heisenbug'' \cite{Heisenbug}, though the more accurate usage is
``Mandelbug'' \cite{GrottkeBugs,FaultTriggers}, since Heisenbugs, properly speaking, are bugs that disappear or
alter behavior under instrumentation (especially in debugging).
Automated test generation systems using test coverage results to drive
the search for interesting inputs may be mislead when a run executes
code only rarely.  And, most significantly, regression testing
effectiveness can be significantly reduced if tests fail only
intermittently and unpredictably, as a result of environmental
factors rather than bugs in changed code.  Such behavior is
unfortunately all too common:  Gao et
al. \cite{Gao:2015:MSU:2818754.2818764} observed coverage differences
of up to 184 lines of code for the same test, and false positive rates
as high as 96\%.  Nondeterminism is sometimes problematic for
developers, who will usually want to  assume that library code they
use behaves in a reliably predictable fashion; nondeterminism is
frequently vexing in debugging efforts; most importantly, perhaps, nondeterminism is \emph{often disastrous} for large-scale highly
automated testing.

@@

A system is deterministic if, given the system's complete state at a point in
time, it is possible, in principle, to predict its future behavior
perfectly.  We say ``in principle'' because in the real world,
prediction may be possible but hopelessly impractical.  We write
complex software systems in many cases because we cannot predict their
behavior (if we could perform the calculations in advance, ourselves,
we would just do so).  As a consequence, in software, rather than
defining determinism in terms of prediction, we usually therefore
simply say that a system is deterministic if, given the same state and
inputs, it always produces the same outputs.

Technically, many ``nondeterministic'' systems are not
nondeterministic in a strong sense, at all.  Given the complete state
of the system (which includes the entire state of the underlying
hardware, the operating system, storage devices, etc.), ignoring
quantum effects, and treating outside interventions such as network
traffic, human activity at an input device, etc. properly as inputs,
most software \emph{is} completely predictable.  What we actually
mean, usually, is that, given a certain limited abstraction of state
and of inputs, observable behavior is repeatable.  This abstraction,
for most software, is not expected to include many elements outside of
the software system itself.

Because the
programmer's abstraction of the system ignores  such a large number of
details, very few tests are in fact ``deterministic'' in the sense
that they eliminate all changes in behavior between executions.  Running the same test
twice almost always results in differences, given a low enough level of
abstraction, since the system load,
cache contents, branch predictor history, etc. are almost never
controlled for; however, this kind of
nondeterminism is usually of no interest, unless the test involves
extremely precise timing constraints.  Rather, what matters is when
some kind of nondeterminism unexpectedly impacts computed values at a
higher level of abstraction; in general, if the operating system
itself is not buggy, low-level nondeterminism, by design, is invisible
except in fine-grained performance testing or real-time systems.
Unexpected nondeterminism usually arises when there is an element of higher-level
state or input that \emph{is} critical to the produced behavior, but
the programmer has not anticipated.  E.g.,  when it is believed that the
behavior of a thread scheduler will not matter, but a race condition
in the code means that it does matter, or when the order of items in
an iterator on a hash table is important, and the hash values used are
randomly salted.

\subsection{The High Cost of Unexpected Nondeterminism}

Unexpected nondeterminism is, unfortunately, usually only discovered
in a context that makes it very hard to debug.   The most common such
contexts are occasional rare failures of a system in deployment, and
regression tests that do not behave reliably (known as flaky tests).
Furthermore, unexpected nondeterminism makes it difficult to use
automated test generation to produce effective regression tests for a
system.  While developers may know how to produce reliably
deterministic unit tests, even in the presence of underlying
nondeterminism, automated test generation, without a large investment
in human time, usually does not have sufficient information to avoid
producing the occasional such test.  Moreover, where a human might
make an assertion about the final value produced during a unit test,
an automatically generated regression test is usually most easily produced by
simply asserting equality with all values produced during a reference
run, since the final step of a test
is likely to be somewhat arbitrary, generated in an effort to increase
code coverage, not establish some functional correctness property.
This may be one reason that automated testing is seldom used to
produce general regression tests.  While failure-inducing tests
produced automatically are often added to regression suites, the full
set of passing tests is, to our knowledge, only infrequently preserved
as part of a typical regression suite, perhaps because such tests are
likely to be \emph{flaky} without considerable human effort.

@@

\subsection{Contributions:  Detecting and Debugging Nondeterminism}

@@

The underlying idea in this paper is that most approaches to
nondeterminism/flaky test detection and debugging rely either on 1) identifying
potential sources of nondeterminism such as test inter-dependence
\cite{LamZE2015} or certain code smells \cite{palomba2017does}  or 2)
use a heuristic approach to mark failures as likely flaky, such as that taken by DeFlaker
\cite{bell2018d}.  In the real world, people usually detect flaky
tests by the most obvious heuristic of all:  they observe a test both
fail and pass for the same code version.  In a sense, even DeFlaker
relies on this method, but avoids having to observe a flaky test actually
passing, by inferring that if a test was previously passing, and
executes no changed code, its failure after a code change \emph{must}
be due to flakiness/nondeterminism.  Our approach to the problem is
fundamentally different, and essentially orthogonal; we concentrate on the detection and avoidance of
nondeterminism/flakiness using \emph{automated test generation},
and make it possible for a developer or test engineer to detect when
important values generated in automated testing differ without a
change in the test itself or the code under test.  While some of these
changes may be harmless, and unable to propagate to cause test
failures, they are potential sources of flaky behavior.  To our
knowledge, this is the first approach to the problem that
allows detection of (potential sources of) flakiness, based on real divergence in
behavior, without having to observe a test actually fail due to nondeterminism.

This approach enables a number of useful results:
 the automated production of
nondeterminism-free regression tests even for libraries with some
sources of nondeterminism, the documentation of sources of
nondeterminism that might produce problems for manually-constructed tests, and, most importantly, detection and effective debugging for
unexpected nondeterminism that is, itself, a bug.  Additonally, by
identifying (using automated testing) library behaviors that are
nondeterministic, our approach should make it much easier to flag
statements in existing human-written unit tests that may produce flaky
behavior:  simply scan the
code for calls to functions identified as sources of nondeterminism.

We use a set
of case studies. including real-world, widely-used, Python libraries to demonstrate the
utility of our ideas.  Failure determinism directly
produces a more than 5\% improvement in mutation score for an already highly
effective automated test generation tool for a file system, with no
additional specification or test design burden.

>>>>

Determinism is particularly important for testing and debugging, where
being able to exactly reproduce system behavior is essential to
productivity \cite{Firesmith}.  If executing ``the same test'' can produce
significantly different behavior each time it is run, the consequences
can be unfortunate.  Developers using a test exhibiting nondeterminism
to debug a system
face a challenge in reasoning about causality, in that an event
observed may not even take place the next time the test is run; this
pernicious phenomenon has been sometimes referred to as the dreaded
``Heisenbug'' \cite{Heisenbug}, though the more accurate usage is
``Mandelbug'' \cite{GrottkeBugs,FaultTriggers}, since Heisenbugs, properly speaking, are bugs that disappear or
alter behavior under instrumentation (especially in debugging).
Automated test generation systems using test coverage results to drive
the search for interesting inputs may go awry when a run executes
code only rarely.  And, most significantly, regression testing
effectiveness can be significantly reduced if tests fail only
intermittently and unpredictably, as a result of environmental
factors rather than bugs in changed code.  Such behavior is
unfortunately all too common:  Gao et
al. \cite{Gao:2015:MSU:2818754.2818764} observed coverage differences
of up to 184 lines of code for the same test, and false positive rates
as high as 96\%.  Nondeterminism is sometimes problematic for
developers, who will usually want to  assume that library code they
use behaves in a reliably predictable fashion; nondeterminism is
frequently vexing in debugging efforts; most importantly, perhaps, nondeterminism is \emph{often disastrous} for large-scale highly
automated testing.

@@

A system is deterministic if, given the system's complete state at a point in
time, it is possible, in principle, to predict its future behavior
perfectly.  We say ``in principle'' because in the real world,
prediction may be possible but hopelessly impractical.  We write
complex software systems in many cases because we cannot predict their
behavior (if we could perform the calculations in advance, ourselves,
we would just do so).  As a consequence, in software, rather than
defining determinism in terms of prediction, we usually 
say that a system is deterministic if, given the same state and
inputs, it always produces the same outputs.

Technically, many ``nondeterministic'' systems are not
nondeterministic in a strong sense, at all.  Given the complete state
of the system (which includes the entire state of the underlying
hardware, the operating system, storage devices, etc.),
most software \emph{is} completely predictable.  What we actually
mean is that, given a certain limited abstraction of state
and inputs, observable behavior is repeatable.  %This abstraction,
%for most software, is not expected to include many elements outside of
%the software system itself.

Because the
programmer's abstraction of the system ignores  such a large number of
details, very few tests are in fact ``deterministic'' in the sense
that they eliminate all changes in behavior between executions.  Running the same test
twice almost always results in differences, given a low enough level of
abstraction, since the system load,
cache contents, branch predictor history, etc. are almost never
controlled for; however, this kind of
nondeterminism is usually of no interest.  Rather, what matters is when
some kind of nondeterminism unexpectedly impacts computed values at a
higher level of abstraction, e.g. when it is believed that the
behavior of a thread scheduler will not matter, but a race condition
in the code means that it does matter, or when the order of items in
an iterator on a hash table is important, and the hash values used are
randomly salted.
\begin{comment} in general, if the operating system
itself is not buggy, low-level nondeterminism, by design, is invisible
except in fine-grained performance testing or real-time systems.
Unexpected nondeterminism usually arises when there is an element of higher-level
state or input that \emph{is} critical to the produced behavior, but
the programmer has not anticipated.  E.g.,  when it is believed that the
behavior of a thread scheduler will not matter, but a race condition
in the code means that it does matter, or when the order of items in
an iterator on a hash table is important, and the hash values used are
randomly salted.
\end{comment}

%\subsection{The High Cost of Unexpected Nondeterminism}

Unexpected nondeterminism is, unfortunately, usually only discovered
in a context that makes it very hard to debug.   The most common such
contexts are occasional rare failures of a system in deployment, and
regression tests that do not behave reliably (known as flaky tests).
Furthermore, unexpected nondeterminism makes it difficult to use
automated test generation to produce effective regression tests for a
system.  While developers may know how to produce reliably
deterministic unit tests, automated test generation usually does not have sufficient information to avoid
producing the occasional such test.  Moreover, where a human might
make an assertion about the final value produced during a unit test,
an automatically generated regression test is usually most easily produced by
asserting equality with all values produced during a reference
run, since the final step of a test
is arbitrary, generated in an effort to increase
code coverage, not establish functional correctness.
This may be one reason that automated testing is seldom used to
produce general regression tests.  While failure-inducing tests
produced automatically are often added to regression suites, the full
set of passing tests is, to our knowledge, only infrequently preserved
as part of a typical regression suite, perhaps because such tests are
likely to be \emph{flaky} without considerable human effort.

@@

\subsection{Contributions}

@@

Previous practical approaches to
nondeterminism/flaky test detection and debugging rely either on 1) identifying
potential sources of nondeterminism such as test inter-dependence
\cite{LamZE2015} or certain code smells \cite{palomba2017does}  or 2)
use a heuristic approach to mark failures as likely flaky, such as that taken by DeFlaker
\cite{bell2018d}.  In the real world, people usually detect flaky
tests by the most obvious heuristic of all:  they observe a test both
fail and pass for the same code version.  In a sense, even DeFlaker
relies on this method, but avoids having to observe a flaky test actually
passing, by inferring that if a test was previously passing, and
executes no changed code, its failure after a code change \emph{must}
be due to flakiness/nondeterminism.  Our approach to the problem is
fundamentally different, and essentially orthogonal; we concentrate on the detection and avoidance of
nondeterminism/flakiness using \emph{automated test generation},
and make it possible for a developer or test engineer to detect when
important values generated in automated testing differ without a
change in the test itself or the code under test.  While some of these
changes may be harmless, and unable to propagate to cause test
failures, they are potential sources of flaky behavior.  To our
knowledge, this is the first approach to the problem that
allows detection of (potential sources of) flakiness, based on real divergence in
behavior, without having to observe a test actually fail due to nondeterminism.

This approach enables a number of useful results:
 the automated production of
nondeterminism-free regression tests even for libraries with some
sources of nondeterminism, the documentation of sources of
nondeterminism that might produce problems for manually-constructed
tests, and, most importantly, detection and \emph{effective debugging}
(using a novel modification of delta-debugging designed for
nondeterminism) for
unexpected nondeterminism that is, itself, a bug.  Additonally, by
identifying library behaviors that are
nondeterministic, our approach should make it much easier to flag
statements in existing human-written unit tests that may produce flaky
behavior:  simply scan the
code for calls to such functions.  We use a set
of case studies. including real-world, widely-used, Python libraries to demonstrate the
utility of our ideas.  Failure determinism directly
produces a more than 5\% improvement in mutation score for an already highly
effective automated test generation tool for a file system, with no
additional specification or test design burden.
